Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2021-09-28 code/MSstatsPTM_Simulation_Analysis.Rmd
Updated main manuscript

## Define metrics
calculate_summary_stats <- function(df, protein_col_name, pval_col_name, sd, rep, conditions){
  fpr <- df %>% filter(get(pval_col_name) < .05 & grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(get(pval_col_name) < .05))
  sensitivity <- df %>% filter(get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(!grepl("NoChange", get(protein_col_name))))
  specificity <- df %>% filter(get(pval_col_name) >= .05 & grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(grepl("NoChange", get(protein_col_name))))
  precision <- df %>% filter(get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(get(pval_col_name) < .05))
  accuracy <- df %>% filter((get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) | 
                              (get(pval_col_name) >= .05 & grepl("NoChange", get(protein_col_name)))) %>% 
    nrow() / nrow(df)
  recall <- df %>% filter(get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / df %>% filter(!grepl("NoChange", get(protein_col_name))) %>% nrow()
  results_temp <- data.table(fpr = fpr, sensitivity = sensitivity, 
                             specificity = specificity, precision = precision, 
                             sd = sd, rep = rep, conditions = conditions,
                             accuracy = accuracy, recall = recall)
  return(results_temp)
}
```

>>>>

## Define metrics
calculate_summary_stats <- function(df, protein_col_name, pval_col_name, sd, rep, conditions){
  fpr <- df %>% filter(get(pval_col_name) < .05 & grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(get(pval_col_name) < .05))
  sensitivity <- df %>% filter(get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(!grepl("NoChange", get(protein_col_name))))
  specificity <- df %>% filter(get(pval_col_name) >= .05 & grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(grepl("NoChange", get(protein_col_name))))
  precision <- df %>% filter(get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / nrow(df %>% filter(get(pval_col_name) < .05))
  accuracy <- df %>% filter((get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) | 
                              (get(pval_col_name) >= .05 & grepl("NoChange", get(protein_col_name)))) %>% 
    nrow() / nrow(df)
  recall <- df %>% filter(get(pval_col_name) < .05 & !grepl("NoChange", get(protein_col_name))) %>% 
    nrow() / df %>% filter(!grepl("NoChange", get(protein_col_name))) %>% nrow()
  results_temp <- data.table(fpr = fpr, sensitivity = sensitivity, 
                             specificity = specificity, precision = precision, 
                             sd = sd, rep = rep, conditions = conditions,
                             accuracy = accuracy, recall = recall)
  return(results_temp)
}

calc_roc <- function(df, protein_col_name, pval_col_name){

  temp_df <- df %>% select(get(protein_col_name), get(pval_col_name))

  temp_df$True <- ifelse(grepl("NoChange", get(protein_col_name)), 0, 1)
  temp_df$Pred <- ifelse(get(pval_col_name) < .05, 1, 0)

}
```

### 2021-09-28 main/MSstatsPTM_Main.tex
Updated main manuscript

\usepackage{fullpage}
\usepackage{amsfonts,amsthm}
\usepackage{amsmath}
\numberwithin{figure}{section} % numbering in each subsection
\numberwithin{table}{section}
\usepackage{setspace}
\usepackage{url}
\usepackage{lscape} %% rotating table
\usepackage{rotating}
\usepackage{hyperref}
\usepackage{authblk}
\usepackage{xcolor} %\[table,dvipsnames]
%\usepackage{makecell}
%\usepackage{array}

@@

\usepackage{xr}
\externaldocument[supp-]{D:\Northeastern\Research\MSstats\MSstatsPTM-manuscript\supplementary\ptm_sm}
%\def\sfigref#1{{\bf Supplementary Fig.~\ref{fig:#1}}}
%\def\snoteref#1{{\bf Supplementary Note~\ref{sec:#1}}}
%\def\snoteshortref#1{{\bf Suppl. Note~\ref{sec:#1}}}
%\def\stabref#1{{\bf Supplementary Table~\ref{tab:#1}}}

@@

The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE. To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method. \cite{satterthwaite:1946} Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., 0.05.\cite{Benjamini:1995} The details are provided in Supplementary Sec. S3.2 \devon{link to supplementary using sfigref}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Analysis with multiple batches}

The proposed statistical framework allows to analyze data from experiments of complex designs such as factorial design. We discuss below a specific design commonly considered in PTM experiments, in which data are acquired in multiple batches, often as a result of multiple repeats for the antibody enrichment step. Typically, the number of spectral features and their intensities are different across batches, and the run-to-run variation sometimes varies across batches as well. The proposed approach summarizes log-intensities in each batch separately, and considers two ways to perform the statistical inference and testing. When variability in site-level summaries and in PTM changes is expected across batches, the proposed approach performs inference of the PTM site abundance and the adjusted difference for each batch separately. The evidence about differential PTM sites between conditions is averaged over batches. Alternatively, assuming identical variability in the site-level summaries and identical differences between conditions across batches (i.e., no interaction effect between condition and batch), the statistical inference and testing can be performed based on one single model with an additional fixed effect of batch. Details about these modeling considerations are discussed in Supplementary \todo{link to Sec. S3.4}.

@@

The statistical modeling approaches discussed above can also be extended to Tandem Mass Tag (TMT) labeling methods. TMT experiments introduce an additional source of variation in the form of different mixtures. To account for the mixture variable, first feature intensities are summarized as described previously, then a new linear mixed effects model with an added term for the mixture is included. This workflow and model follow the methods used for MSstatsTMT. \cite{Huang:2020} These methods are repeated to quantify and model global protein abundance. Once both PTM and global protein are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods are before.

\todo{add figref for TMT model}
\todo{add figref for TMT experiment table}

@@

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test or limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverage measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Details of the simulations can be found in Supplementary %\cref{supp-sec:sim}.

@@

In this simulation, a clean dataset was generated with a high number of modified and unmodified peptide features and no missing values. Not accounting for changes in unmodified peptide level resulted in a high false discovery rate for all methods\todo{Reference supplementary figure}. When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR. When looking at recall and accuracy the proposed approach and limma performed similarly, while t-test lagged behind. Two-sample t-test only used data within the groups of interest while ignoring the rest of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach and limma leveraged all available information, which resulted in improved power with increased number of conditions. Further results are provided in Supplementary \todo{link to Sec. S4.1.1}.

@@

In this simulation real world experimental conditions were introduced, including missing values and low number of modified features. In this case, changes in unmodified peptide level still needed to be accounted for to control the FDR in all methods. Once controlling for changes in unmodified peptide, it is clear that the proposed method performed stronger across all methods when compared to limma and t-test. The proposed method well calibrated the Recall rate and outperformed the other methods. Additionally, in overall accuracy the proposed method showed the strongest performance, even when the number of biological replicates were lower. Finally, the distribution of estimated fold changes is shown in \todo{reference sim2 boxplot}. The proposed method shows a tighter distribution of estimated fold changes around the true fold change, while the other methods showed much wider distributions. Further results are provided in Supplementary \todo{link to Sec. S4.1.2}.

@@

We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark. Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions. Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change. \todo{Add figure of mixtures} Two sets of data were acquired for each mixture: modifed peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate. These datasets were used as the PTM and global protein data.

\todo{figref volcano plot of spike in peptides}

In *FigXX* we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level. After adjustment the estimated fold change is more in line with expectation. Additionally, background peptides serving as the null model, colored grey, show many false positives before adjustment is made. Again after adjustment the results improve and the number of false positives decrease significantly.

\todo{Add limma plot}

In *FigXX* we can see the results of the same experiment modeled using limma with protein adjustment discussed in Section~\ref{sec:comp_sim}. Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied. However, using limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue. In this case, usin limma would have resulted in missing the majority of differential modified peptides.  In terms of false positives, the results are very similar to the propsed method, with many false positives before adjustment and much fewer afterward.

Further results of this experiment are provided in Supplementary \todo{link to Sec. S4.2}.

@@

The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was inbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.

Looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method. The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This is contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e^{-8}$. In this case the effect of the modified peptide is strongly confounded with changes in the global protein. The proposed method allows us to remove this confounding and estimate the true effect. The results of this can be seen in Figure \todo{add profile plot}.

Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.1}.

@@

Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.2}.

@@

This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 overexpression (USP30 OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides. A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary.

@@

Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.3}.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/fig2.png}
\caption{\todo{todo}}
\label{fig:batch}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/fig3.png}
\caption{
Data structure of a typical PTM experiment and goals of PTM characterization.(a)Schematic data representation, in a simplified case of two conditions and two replicate runs. Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i. Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by *i. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). Some spectral features can be missing.(b)PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions. \todo{there is no d..}(d) Statistical experimental design in terms of sample size calculations and power analysis.}
\label{fig:data-structure}
\end{figure}

{\bf Figure 4} False positive rate and statistical power in PTM significance analysis by the proposed approach, t-test with protein-level adjustment, and t-test without protein-level adjustment. (a) When there was no change in protein abundance, all considered methods well calibrated the Type I error rate. (b) When the PTM changes were entirely due to changes in protein abundance across conditions, analysis accounting for the protein-level changes resulted in off-target, high false positive rates. (c) Statistical powers by the proposed approach and t-test with protein-level adjustment, where the data consisted of 2 conditions and the SD corresponding unmodified feature intensities was 0.2. (d) Statistical powers by the proposed approach and t-test with protein-level adjustment, where the data consisted of 2 conditions, the SD corresponding unmodified feature intensities was 0.2, and the PTM was missing in Run 1 of Condition 1. (e) Same as in (d), but the data consisted of 4 conditions.

{\bf Figure 5} Results corresponding to estimation error and false positive rate and statistical power of the PTM significance analysis, where the data were acquired in two batches. The following parameters were considered to represent the batch effects: no batch-condition interaction, mean intensity level in Batch 2 was higher than Batch 1 by 2 on log scale, and the SDs in Batch 1 and Batch 2 were 0.2 and 0.3, respectively.(a)Estimation based on the most statistically significant batch with t-test was more variable than other methods and frequently biased. (b)The proposed approach better calibrated Type I error rate. (c)The proposed approach improved statistical power in all the considered cases with different number of conditions and replicates. This was achieved by properly characterizing batch effects and leveraging all available information. Ignoring batch effects by t-test (no batch) lost power dramatically. The negative impact was only partially reduced by increasing the sample size to 5.

{\bf Figure 6} Design of future PTM experiments in terms of sample size calculations and power analysis. (a) Protein-level adjustment relies on the inference of protein abundance, which introduces additional uncertainty in the estimate of PTM difference. Therefore, the required sample size to detect a systematic change is higher than as expected for standard differential analysis without adjustment. Sample size calculations without accounting for the uncertainty would lead to over-optimistic, under-powered studies. (b) In complex designs, simultaneously analyzing all the conditions effectively increases the degrees of freedom and requires fewer replicates. (c)Increasing the sample size and analyzing multiple conditions together both result in improved statistical power.

>>>>

\usepackage{fullpage}
\usepackage{amsfonts,amsthm}
\usepackage{amsmath}
\numberwithin{figure}{section} % numbering in each subsection
\numberwithin{table}{section}
\usepackage{setspace}
\usepackage{url}
\usepackage{lscape} %% rotating table
\usepackage{rotating}
\usepackage{authblk}
\usepackage{xcolor} %\[table,dvipsnames]
%\usepackage{makecell}
%\usepackage{array}

@@

\usepackage{xr}
\externaldocument[supp-]{../supplementary/ptm_sm}
%\def\sfigref#1{{\bf Supplementary Fig.~\ref{fig:#1}}}
%\def\snoteref#1{{\bf Supplementary Note~\ref{sec:#1}}}
%\def\snoteshortref#1{{\bf Suppl. Note~\ref{sec:#1}}}
%\def\stabref#1{{\bf Supplementary Table~\ref{tab:#1}}}

@@

The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE. To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method. \cite{satterthwaite:1946} Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$.\cite{Benjamini:1995} More details are provided in Supplementary Sec \ref{supp-sec:test}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection*{Analysis with multiple batches}

%The proposed statistical framework allows to analyze data from experiments of complex designs such as factorial design. We discuss below a specific design commonly considered in PTM experiments, in which data are acquired in multiple batches, often as a result of multiple repeats for the antibody enrichment step. Typically, the number of spectral features and their intensities are different across batches, and the run-to-run variation sometimes varies across batches as well. The proposed approach summarizes log-intensities in each batch separately, and considers two ways to perform the statistical inference and testing. When variability in site-level summaries and in PTM changes is expected across batches, the proposed approach performs inference of the PTM site abundance and the adjusted difference for each batch separately. The evidence about differential PTM sites between conditions is averaged over batches. Alternatively, assuming identical variability in the site-level summaries and identical differences between conditions across batches (i.e., no interaction effect between condition and batch), the statistical inference and testing can be performed based on one single model with an additional fixed effect of batch. Details about these modeling considerations are discussed in Supplementary Sec \ref{supp-sec:test}.

@@

The statistical modeling approaches discussed above can also be extended to Tandem Mass Tag (TMT) labeling methods. TMT experiments introduce an additional source of variation in the form of different mixtures. To account for mixture variance, first feature intensities are summarized again using Tukey's Median Polish and missing values are imputed, if desired, then a new linear mixed effects model with an added term for the mixture is included. The workflow and statisical model described generally follow the methods used for modeling proteins in MSstatsTMT. \cite{Huang:2020} These methods are repeated to quantify both modified and unmodified peptide abundance. Once both modified and unmodified peptides are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods described previously.

For more information about the extension to Tandem Mass Tag experiments see Supplementary Sec. \ref{supp-sec:tmtmethod}.

@@

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test or limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverage measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Details of the results of this simulation can be found in Supplementary Sec. \ref{supp-sec:sim}.

@@

In this simulation, a clean dataset was generated with a high number of modified and unmodified peptide features and no missing values. Not accounting for changes in unmodified peptide level resulted in a high false discovery rate for all methods, seen in Figure \ref{fig:fdr_all_models}. When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR. When looking at recall and accuracy the proposed approach and limma performed similarly, while t-test lagged behind. Two-sample t-test only used data within the groups of interest while ignoring the rest of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach and limma leveraged all available information, which resulted in improved power with increased number of conditions. Further results are provided in Supplementary Figures \ref{supp-fig:sim1_fpr} to \ref{supp-fig:sim1_recall}.

@@

In this simulation real world experimental conditions were introduced, including missing values and low number of modified features. In this case, changes in unmodified peptide level still needed to be accounted for to control the FDR in all methods. Once controlling for changes in unmodified peptide, it is clear that the proposed method performed stronger across all methods when compared to limma and t-test, as seen in Figure \ref{fig:sim_accuracy}. The proposed method well calibrated the Recall rate and outperformed the other methods. Additionally, in overall accuracy the proposed method showed the strongest performance, even when the number of biological replicates were lower. Finally, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, while the other methods showed much wider distributions. Further results are provided in Supplementary Figures \ref{supp-fig:sim2_recall} to \ref{supp-fig:fc_boxplot}.

@@

We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark. Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions. Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change. Two sets of data were acquired for each mixture: modified peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate. These datasets were used as the PTM and global protein data.

In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level. After adjustment the estimated fold change is more in line with expectation. Additionally, background peptides serving as the null model, colored grey, show many false positives before adjustment is made. Again after adjustment the results improve and the number of false positives decrease significantly.

In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using limma with protein adjustment discussed in Section~\ref{sec:comp_sim}. Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied. However, using limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue. In this case, using limma would have resulted in missing the majority of differential modified peptides.  In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer afterward.

Further results of this experiment are provided in Supplementary Sec. \ref{supp-sec:benchmark}.

@@

The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.

Looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method. The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This is contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e^{-8}$. In this case the effect of the modified peptide is strongly confounded with changes in the global protein. The proposed method allows us to remove this confounding and estimate the true effect. The results of this can be seen in Figure \ref{fig:ipah_profile}.

Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:ipah}.

@@

Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.

@@

This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides. A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary.

@@

Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:usp30}.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/fig3.png}
\caption{
Data structure of a typical PTM experiment and goals of PTM characterization. (a) Schematic data representation, in a simplified case of two conditions and two replicate runs. Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i. Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by i*. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). Some spectral features can be missing. (b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{All the considered methods in simulation 1 correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
}\label{fig:fdr_all_models}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{The advantage of using the proposed approach is apparent when looking at simulation 2, which includes limited observations and the presence of missing values. In overall accuracy, the proposed method performs stronger than limma and t-test in nearly every model. Even at lower replicates the proposed method still outperformed limma. The lowest performaning method was $t$-test. Limma shows comparable performance to the proposed method in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
\label{fig:sim_accuracy}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/spike_in_msstatsptm_volcano.png}
\caption{Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment. After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation. Additionally the background grey colored peptides showed many false positives before adjustment. After adjustment the false positives were decreased considerably.
}\label{fig:spikein_prop_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
\caption{Limma method used to model the spike in experiment. Using limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level. However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue. In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer after.
}\label{fig:spikein_limma_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
\caption{Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$. The individual PSM features are shown in grey, while the feature summarization is shown in red. When looking at the summary of the modification and global protein it is clear the conditions follow different trends. Specifically, there appears to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there is a large negative change when looking at the unmodified plot. This indicates the modification is confounded with changes in the unmodified protein.
}\label{fig:ipah_profile}
\end{figure}

%{\bf Figure 4} False positive rate and statistical power in PTM significance analysis by the proposed approach, t-test with protein-level adjustment, and t-test without protein-level adjustment. (a) When there was no change in protein abundance, all considered methods well calibrated the Type I error rate. (b) When the PTM changes were entirely due to changes in protein abundance across conditions, analysis accounting for the protein-level changes resulted in off-target, high false positive rates. (c) Statistical powers by the proposed approach and t-test with protein-level adjustment, where the data consisted of 2 conditions and the SD corresponding unmodified feature intensities was 0.2. (d) Statistical powers by the proposed approach and t-test with protein-level adjustment, where the data consisted of 2 conditions, the SD corresponding unmodified feature intensities was 0.2, and the PTM was missing in Run 1 of Condition 1. (e) Same as in (d), but the data consisted of 4 conditions.

%{\bf Figure 5} Results corresponding to estimation error and false positive rate and statistical power of the PTM significance analysis, where the data were acquired in two batches. The following parameters were considered to represent the batch effects: no batch-condition interaction, mean intensity level in Batch 2 was higher than Batch 1 by 2 on log scale, and the SDs in Batch 1 and Batch 2 were 0.2 and 0.3, respectively.(a)Estimation based on the most statistically significant batch with t-test was more variable than other methods and frequently biased. (b)The proposed approach better calibrated Type I error rate. (c)The proposed approach improved statistical power in all the considered cases with different number of conditions and replicates. This was achieved by properly characterizing batch effects and leveraging all available information. Ignoring batch effects by t-test (no batch) lost power dramatically. The negative impact was only partially reduced by increasing the sample size to 5.

%{\bf Figure 6} Design of future PTM experiments in terms of sample size calculations and power analysis. (a) Protein-level adjustment relies on the inference of protein abundance, which introduces additional uncertainty in the estimate of PTM difference. Therefore, the required sample size to detect a systematic change is higher than as expected for standard differential analysis without adjustment. Sample size calculations without accounting for the uncertainty would lead to over-optimistic, under-powered studies. (b) In complex designs, simultaneously analyzing all the conditions effectively increases the degrees of freedom and requires fewer replicates. (c)Increasing the sample size and analyzing multiple conditions together both result in improved statistical power.
